Fast sparse period estimation

### 2014-07-29 paper.tex
minor

The problem of estimating the period of a point process from observations that are both sparse and noisy is considered.  By sparse it is meant that only a potentially small unknown subset of the process is observed.  By noisy it is meant that the subset that is observed, is observed with some error, or noise.  Existing accurate algorithms for estimating the period require $O(N^2)$ operations where $N$ is the number of observations.  By quantizing the observations we produce an estimator that requires only $O(N\log N)$ operations by use of the chirp z-transform or the fast Fourier transform.  The quantization has the adverse effect of decreasing the accuracy of the estimator.  This is investigated by Monte-Carlo simulation.  The simulations indicate that significant computational savings are possible with negligible loss in statistical accuracy.

@@

In Section~\ref{sec:quant-peri} we show that, by quantizing the observations $y_1,\dots,y_N$, an approximation to the periodogram estimator can be computed using only $O(N\log N)$ operations.  This is facilitated by either the chirp z-transform~\cite{Rabiner1969} or the fast Fourier transform.  The quantization has the adverse affect of decreasing the accuracy of the estimator.  The estimator is accurate if a sufficiently fine quantizer is chosen, but this requires a larger transform.  The trade off between quantization level, accuracy, and computational complexity is investigated by computer simulation in Section~\ref{sec:simulations}.  The simulations indicate that significant computational savings are possible with negligible loss in accuracy.  Sadler and Casey~\cite{726812} have previously suggested use of the fast Fourier transform for period estimation.  However, their estimator requires larger transforms and produces less accurate results than the estimator described here (see Sec.~VI and Fig. 1-4 in~\cite{726812}).  %Our estimator has accuracy similar to the least squares estimator, i.e., similar to the maximum likelihood estimator if the noise $w_1,\dots,w_N$ is assumed independent and identically normally distributed.

@@

Denote by $\round{x} = \sfloor{x + \tfrac{1}{2}}$ the nearest integer to $x$.  %The direction of rounding for half integers is not important.  Half integers are rounded up in our implementation.  
Let $q$ be a positive real number and define the quantization function
\[
Q(x) = \round{qx}/q
\]
that rounds its argument to a nearest multiple of $1/q$.  Large $q$ corresponds with fine quantization and small $q$ corresponds with coarse quantization.  Compute quantized observations
\[
z_n = Q(y_n) =  \ell_n/q, \qquad n = 1,\dots,N
\] 
where $\ell_n = \round{qy_n} \in \ints$.  The periodogram of the quantized observations $z_1,\dots,z_N$ is
\begin{align*}
I_z(f) = \babs{ \sum_{n=1}^N e^{ -2\pi j f z_n} }^2 = \babs{ \sum_{n=1}^N e^{ -2\pi j f \ell_n/q} }^2.
\end{align*}
Observe that $I_z$ is periodic with period $q$ and so quantization should be fine enough that 
\begin{equation}\label{eq:boundaliasingP}
q > f_{\text{max}} - f_{\text{min}}.
\end{equation}
Otherwise there would exist frequencies in the interval $(f_{\text{min}}, f_{\text{max}})$ that are aliases with respect to $I_z$.  %That is, there would exists two frequences $f_1,f_2 \in (f_{\text{min}}, f_{\text{max}})$ such that $f_1 = f_2 + 1/P$ and $I_z(f_1) = I_z(f_2)$.

@@

The previous section shows how the approximate maximiser $\widetilde{f}^\prime$ can be computed using either the chirp z-transform or a fast Fourier transform of length $M$.  We now describe recommended values for the parameters $q, M$, and $D$.  The simulations presented in Section~\ref{sec:simulations} suggest that an effective value for the quantization parameter is $q = 5 f_{\text{max}}$.  If the chirp z-transform is used then $D$ is chosen according to~\eqref{eq:gridwidthrecommended}.  The chirp z-transform requires computation of three fast Fourier transforms, each of length $L + K - 1$, where 
\[
L = \ellmax - \ellmin + 1 = \round{q\ymax} - \round{q\ymin} + 1.
\]
If instead a single fast Fourier transform is used then our recommended value for the length of the transform is $M= 2L$.  In this case the grid distance
\[ 
D = \frac{q}{M} = \frac{q}{2(\round{q\ymax} - \round{q\ymax} + 1)} \approx  \frac{1}{2(\ymax - \ymin)}
\]
is approximately equal to~\eqref{eq:gridwidthrecommended}.

>>>>

The problem of estimating the period of a point process from observations that are both sparse and noisy is considered.  By sparse it is meant that only a potentially small unknown subset of the process is observed.  By noisy it is meant that the subset that is observed, is observed with error, or noise.  Existing accurate algorithms for estimating the period require $O(N^2)$ operations where $N$ is the number of observations.  By quantizing the observations we produce an estimator that requires only $O(N\log N)$ operations by use of the chirp z-transform or the fast Fourier transform.  The quantization has the adverse effect of decreasing the accuracy of the estimator.  This is investigated by Monte-Carlo simulation.  The simulations indicate that significant computational savings are possible with negligible loss in statistical accuracy.

@@

In Section~\ref{sec:quant-peri} we show that, by quantizing the observations $y_1,\dots,y_N$, an approximation to the periodogram estimator can be computed using only $O(N\log N)$ operations.  This is facilitated by either the chirp z-transform~\cite{Rabiner1969} or the fast Fourier transform.  The quantization has the adverse affect of decreasing the accuracy of the estimator.  The estimator is accurate if a sufficiently fine quantizer is chosen, but this requires a larger transform.  The trade off between quantization level, accuracy, and computational complexity is investigated by computer simulation in Section~\ref{sec:simulations}.  The simulations indicate that significant computational savings are possible with negligible loss in accuracy.  Sadler and Casey~\cite{726812} have previously suggested use of the fast Fourier transform for period estimation.  However, their estimator requires larger transforms and produces less accurate results than the estimator described here (see Sec.~VI in~~\cite{726812} and Fig. 1-4 in~\cite{726812}).  %Our estimator has accuracy similar to the least squares estimator, i.e., similar to the maximum likelihood estimator if the noise $w_1,\dots,w_N$ is assumed independent and identically normally distributed.

@@

Denote by $\round{x} = \sfloor{x + \tfrac{1}{2}}$ the nearest integer to $x$.  %The direction of rounding for half integers is not important.  Half integers are rounded up in our implementation.  
Let $q$ be a positive real number and define the quantization function
\[
Q(x) = \round{qx}/q
\]
that rounds its argument to a nearest multiple of $1/q$.  Large $q$ corresponds with fine quantization and small $q$ corresponds with coarse quantization.  Compute quantized observations
\[
z_n = Q(y_n) =  \ell_n/q, \qquad n = 1,\dots,N
\] 
where $\ell_n = \round{qy_n} \in \ints$.  The periodogram of the quantized observations $z_1,\dots,z_N$ is
\begin{align*}
I_z(f) = \babs{ \sum_{n=1}^N e^{ -2\pi j f z_n} }^2 = \babs{ \sum_{n=1}^N e^{ -2\pi j f \ell_n/q} }^2.
\end{align*}
Observe that $I_z$ is periodic with period $q$ and so quantization must be fine enough that 
\begin{equation}\label{eq:boundaliasingP}
q > f_{\text{max}} - f_{\text{min}}.
\end{equation}
Otherwise there would exist frequencies in the interval $(f_{\text{min}}, f_{\text{max}})$ that are aliases with respect to $I_z$.  %That is, there would exists two frequences $f_1,f_2 \in (f_{\text{min}}, f_{\text{max}})$ such that $f_1 = f_2 + 1/P$ and $I_z(f_1) = I_z(f_2)$.

@@

The previous section shows how the approximate maximiser $\widetilde{f}^\prime$ can be computed using either the chirp z-transform or a fast Fourier transform of length $M$.  We now describe recommended values for the parameters $q$, $M$, and $D$.  The simulations presented in Section~\ref{sec:simulations} suggest that an effective value for the quantization parameter is $q = 5 f_{\text{max}}$.  If the chirp z-transform is used then $D$ is chosen according to~\eqref{eq:gridwidthrecommended}.  The chirp z-transform requires computation of three fast Fourier transforms, each of length $L + K - 1$, where 
\[
L = \ellmax - \ellmin + 1 = \round{q\ymax} - \round{q\ymin} + 1.
\]
If instead a single fast Fourier transform is used then our recommended value for the length of the transform is $M= 2L$.  In this case the grid distance
\[ 
D = \frac{q}{M} = \frac{q}{2(\round{q\ymax} - \round{q\ymax} + 1)} \approx  \frac{1}{2(\ymax - \ymin)}
\]
is approximately equal to~\eqref{eq:gridwidthrecommended}.
